---
title: "Breast cancer proteomics data analysis [Brigham & Women's Hospital & Harvard Medical School (2012)]"
format: html
editor: visual
---

## Useful libraries

```{r, useful-libraries}
library(tidyverse)
library(normalblockr)
library(aricode)
library(clusterProfiler)
library(pathview)
library(enrichplot)
set.seed(1)
```

## Loading & formatting data

Loading clinical data to get cancer subtypes as covariates.

```{r, loading-clinical-data}
clinical  <- read_rds("brca_clinical.rds")
```

Loading proteomics data to analyse.

```{r}
RPPA_expr <- read_rds("RPPA_expr.rds")
RPPA_meta <- read_rds("RPPA_meta.rds")
RPPA_clin <- clinical[match(rownames(RPPA_expr), clinical$sampleId), ]
```

Formatting data for Normal-Block.

```{r, Normal-Block-formatting}
Y            <- as.matrix(RPPA_expr) %>% scale()
X_subtype    <- model.matrix(~ 0 + PAM50_SUBTYPE, data = RPPA_clin)
data_subtype <- normal_data$new(Y, X_subtype)
```

## Applying Normal-Block to the data

### Running the integrated inference procedure for 1 to 50 clusters.

```{r, running-normal-block}
NB_prot_subtype <- normal_block(data_subtype, blocks = 1:50)
```

### Model selection: criteria to fix the number of clusters

<<<<<<< HEAD
```{r, plotting-criteria}
NB_prot_subtype$plot(c("BIC", "ICL"))
=======
```{r, plotting criteria}
NB_prot_subtype$plot(c("deviance", "BIC", "ICL"))
>>>>>>> b1aa640e96852ce025a4e129721f061a1b3c82b7
```

```{r, model-selection}
selected_NB <- NB_prot_subtype$get_best_model("ICL")
paste0("ICL selects ", selected_NB$Q, " clusters.")
```

## Enrichment analysis

```{r, enrichment}
EntrezGeneID <- str_remove_all(RPPA_meta["entrezGeneId", ], " ")
df_inter <- data.frame(Entrez = EntrezGeneID, group  = selected_NB$clustering)
inter_kegg <- compareCluster(Entrez~group, data = df_inter, fun='enrichKEGG', pvalueCutoff = 0.001)
inter_kegg <- setReadable(inter_kegg, OrgDb = "org.Hs.eg.db", keyType="ENTREZID")
dotplot(inter_kegg, font.size = 8)
```
