---
title: "normalblockr: Gaussian graphical models with latent clustering structure for multivariate continuous data"
format:
  gfm:
    default-image-extension: ""
---

<!-- badges: start -->
[![R-CMD-check](https://github.com/jchiquet/normalblockr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jchiquet/normalblockr/actions/workflows/R-CMD-check.yaml)
[![Coverage status](https://codecov.io/gh/jchiquet/normalblockr/branch/master/graph/badge.svg)](https://app.codecov.io/github/jchiquet/normalblockr?branch=master)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html)
[![License: GPL v3](https://img.shields.io/badge/License-GPLv3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
[![](https://img.shields.io/github/last-commit/jchiquet/normalblockr.svg)](https://github.com/jchiquet/normalblockr/commits/master)
<!-- badges: end -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  cache      = TRUE,
  cache.lazy = FALSE, # R6 model objects don't survive knitr's lazy-load cache promises
  # ... and a cached R6 object is only as good as the class definition that
  # made it: a chunk whose code hasn't changed is restored as-is, so an object
  # built by an older normalblockr keeps that version's methods while the
  # chunks around it run against the new one. Keying the cache on the
  # installed package makes every reinstall invalidate it.
  cache.extra = file.mtime(system.file("Meta", "package.rds", package = "normalblockr")),
  fig.path   = "man/figures/README-",
  fig.width  = 7, fig.height = 5
)
```

## Description

> The Normal-Block model[^1] is a Gaussian graphical model with a latent clustering structure, designed for the multivariate analysis of continuous data: it clusters variables and, building on the graphical lasso, infers a network of statistical dependencies *between clusters* rather than between individual variables. This package implements an efficient (variational) EM algorithm to fit it, accompanied by a set of functions for model selection, visualization and diagnostic. See [all the dedicated vignettes](https://jchiquet.github.io/normalblockr/articles/) for a comprehensive introduction.

**normalblockr** covers the following model variants, all built around the same `normal_block()`/`NormalBlockData` interface:

- **Known clusters**: the grouping of variables is given (e.g. from prior knowledge); only the association network between clusters is estimated, by EM.
- **Unknown clusters**: the grouping is inferred jointly with everything else by a variational EM, either for a single number of clusters or over a range explored as a collection, with model selection via BIC/EBIC/ICL.
- **Sparse (graphical-lasso) network**: an $\ell_1$ penalty[^2] on the inter-cluster precision matrix, for a single penalty value or a path explored as a collection.
- **Zero-inflated extension**: an excess-of-zeros layer for data (e.g. abundance/biomass tables) with more exact zeros than a plain Normal model can represent.

Any combination of these is reached through the same `normal_block()` function: known or unknown clustering, sparse or not, zero-inflated or not are independent choices, not separate model classes to learn.

Since version 0.3.0 the package also fits a second, complementary family, reached with `normal_block(..., model = "mean")`:

- **Mean-block models**: variables are clustered by how they *respond* to the covariates rather than by how they *covary*. All variables in a cluster share one regression profile, so the mean is $\mu_i = C B^\top X_i$ with $B$ of size $d \times q$ instead of $d \times p$. Known or unknown clustering, zero-inflated or not, with the residual covariance taken diagonal (the default), spherical, or full.

The two families answer different questions and generally return different groupings; `normal_block_sequential()` runs one after the other when both are of interest. See the [mean-block vignette](https://jchiquet.github.io/normalblockr/articles/mean-block-breast-cancer.html).

[^1]: Tous, J., & Chiquet, J. (2026). An integrated method for clustering and association network inference. Computational Statistics & Data Analysis, 219, 108347. [doi:10.1016/j.csda.2026.108347](https://doi.org/10.1016/j.csda.2026.108347)
[^2]: Friedman, J., Hastie, T. and Tibshirani, R. Sparse inverse covariance estimation with the graphical lasso. Biostatistics, 9(3), 2008.

## Installation

Install the released version from CRAN:

```{r package-cran, eval = FALSE}
install.packages("normalblockr")
```

or the development version from [GitHub](https://github.com/jchiquet/normalblockr):

```{r package-github, eval = FALSE}
# install.packages("pak")
pak::pak("jchiquet/normalblockr")
```

## Illustration

We illustrate the known-/unknown-clusters and sparse variants on `brca_rppa`[^3]: reverse-phase protein array measurements of 163 proteins across 346 breast cancer tumor samples from The Cancer Genome Atlas, together with each sample's PAM50 molecular subtype.

[^3]: Cancer Genome Atlas Network. Comprehensive molecular portraits of human breast tumours. Nature, 490, 2012, 61-70. [doi:10.1038/nature11412](https://doi.org/10.1038/nature11412)

```{r load}
library(normalblockr)
data(brca_rppa)
Y    <- as.matrix(brca_rppa$expr)
X    <- model.matrix(~ 0 + PAM50_SUBTYPE, data = brca_rppa$covariates)
data <- NormalBlockData$new(Y, X)
```

### Known clusters

A simple, fully data-driven grouping (hierarchical clustering on the raw expression profiles, with no reference to the Normal-Block model) handed to `normal_block()` as fixed -- only the network between the 6 blocks is estimated.

```{r known}
hc    <- brca_rppa$expr |> scale() |> t() |> dist() |> hclust("ward.D2")
group <- cutree(hc, 6) |> normalblockr:::as_indicator()
m_known <- normal_block(data, blocks = group)
m_known
```

### Unknown number of clusters

Leave the clustering for the model to infer, over a range of candidate cluster counts explored as a collection, then select by ICL.

```{r unknown}
m_unknown <- normal_block(data, blocks = 2:20)
m_unknown$refine(verbose = FALSE)
m_unknown$plot(c("deviance", "BIC", "ICL", "EBIC"))
```

```{r unknown-best}
m_unknown$get_best_model("EBIC")
```

### Sparse network

Treating a clustering as fixed (known, or selected above), explore a path of graphical-lasso penalties on the inter-cluster network and select by BIC. Unlike the dense, unpenalized network, this is the one actually worth visualizing.

```{r sparse}
m_sparse <- normal_block(data, blocks = group, sparsity = TRUE)
sp_best  <- m_sparse$get_best_model("BIC")
sp_best$plot_network()
```

### Zero-inflated data

For data with an excess of exact zeros beyond what a plain Normal model would represent (e.g. abundance/biomass tables), `zero_inflation = TRUE` adds a per-variable excess-of-zero layer. Illustrated on `onema`[^4]: total biomass of 46 fish species across 399 French stream electrofishing stations, where seven in ten entries are exact zeros.

[^4]: Danet, A., Mouchet, M., Bonnaffé, W., Thébault, E., Fontaine, C. Species richness and food-web structure jointly drive community biomass and its temporal stability in fish communities, data set, 2021, Zenodo. [doi:10.5281/zenodo.5095656](https://doi.org/10.5281/zenodo.5095656)

```{r zi}
data(onema)
X_zi    <- model.matrix(~ 1 + temperature_med, data = onema$covariates)
Y_zi    <- log(1 + onema$biomass)
data_zi <- NormalBlockData$new(Y_zi, X_zi)
m_zi    <- normal_block(data_zi, blocks = 2:8, zero_inflation = TRUE)
m_zi$get_best_model("ICL")
```

## Learning more

- [`normal-block`](https://jchiquet.github.io/normalblockr/articles/normal-block.html) (`vignette("normal-block")`): a general introduction on simulated data -- known/unknown clusters, sparsity, zero-inflation.
- [`zero-inflated-normal-block`](https://jchiquet.github.io/normalblockr/articles/zero-inflated-normal-block.html) (`vignette("zero-inflated-normal-block")`): a full worked example on `onema`, with the math behind the zero-inflation extension.
- [`breast-cancer-proteomics`](https://jchiquet.github.io/normalblockr/articles/breast-cancer-proteomics.html) (`vignette("breast-cancer-proteomics")`): a full worked example on `brca_rppa` -- known vs. inferred clustering, model selection, the post-hoc `refine()` step, sparsifying the selected network.
- [`mean-block-breast-cancer`](https://jchiquet.github.io/normalblockr/articles/mean-block-breast-cancer.html) (`vignette("mean-block-breast-cancer")`): the mean-block family on `brca_rppa` -- clustering proteins by their subtype signature, choosing the shape of the residual covariance, sparsifying it.
- `inst/normal_block_models.qmd` in the package sources: a reference card in two parts, the models (both families, with their criteria and E/M or VE/M updates) and the implementation notes.

## References
